---
output:
  html_document: default
  pdf_document: default
---
# Practical Machine Learning Course Project

#### By Khew Kok Seong

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement Â- a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks.

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this data set, the participants were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

In this report we aim to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants toto predict the manner in which praticipants did the exercise.  This report covers:-     
   * how the model is built   
   * use of cross validation   
   * an estimate of expected out of sample error   
   
The dependent variable is the `classe` variable in the training set. 

## Getting and cleaning the Data
Assume all the required packages are pre-installed, proceed to load the libraries and set the seed for reproduceability:
```{r error=FALSE, warning=FALSE, message=FALSE}
library(caret)
library(randomForest) 
library(rpart)
library(rpart.plot)
library(e1071)

set.seed(1375)
```

Load the data sets into R and make sure that missing values are coded correctly.
Irrelevant variables will be deleted.
Results will be hidden from the report for clarity and space considerations.
```{r warning=FALSE}
# Assume both data sets are saved into the working directory, in my case, it is "C:/tmp2/coursera"
# Some missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.
# We notice that both data sets contain columns with all missing values - these will be deleted.  

# Loading the training data set into my R session replacing all missing with "NA"
training <- read.csv("C:/tmp2/coursera/pml-training.csv", na.strings=c("NA","#DIV/0!", ""))

# Loading the testing data set 
testing <- read.csv('C:/tmp2/coursera/pml-testing.csv', na.strings=c("NA","#DIV/0!", ""))

# Check dimensions for number of variables and number of observations
dim(training)
dim(testing)

# Delete columns with all missing values
training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]
```

```{r}
# Some variables are irrelevant to our current project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7). We can delete these variables.
training <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]

```
      
       
## Partitioning the training data set to allow cross-validation
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (70%) and subTest (30%).
This will be performed using random subsampling without replacement.
```{r}
subsamples <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
subTraining <- training[subsamples, ] 
subTesting <- training[-subsamples, ]
dim(subTraining)
dim(subTesting)
```
     
     
### First prediction model: Using Decision Tree
```{r}
dt_model <- rpart(classe ~ ., data=subTraining, method="class")

# Predicting:
prediction1 <- predict(dt_model, subTesting, type = "class")

# Plot of the Decision Tree
rpart.plot(dt_model, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```


```{r}
# Test results on our subTesting data set:
confusionMatrix(prediction1, subTesting$classe)
```
    
    
### Second prediction model: Using Random Forest
```{r}
rf_model <- randomForest(classe ~. , data=subTraining, method="class")

# Predicting:
prediction2 <- predict(rf_model, subTesting, type = "class")

# Test results on subTesting data set:
confusionMatrix(prediction2, subTesting$classe)
```
   
   
### Decision
The random Forest model is choosen because of the accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to 0.745 (95% CI: (0.734, 0.756)) for Decision Tree model. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as `1 - accuracy` for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we are confident that very few, or none, of the test samples will be missclassified.


## Submission
```{r}
# predict outcome levels on the original Testing data set using Random Forest algorithm
predictfinal <- predict(rf_model, testing, type="class")
predictfinal
```
     
```{r}
# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)
```